Make any stupid plot you want

### 2021-02-16 draft/sections/intro.tex
notation call out + start of math revisions
rework of data definitions

As described by Mackinlay, a visualization tool produces a graphical design and an image rendered based on that design. He defines the graphical design as the set encoding relations from data to visual representation\cite{mackinlayAutomatingDesignGraphical1986}, and the design rendered in an idealized abstract space is what throughout this paper we will refer to as a graphic. Mackinlay  proposes that a visualization tool's expressiveness is a measure of how much of the structure of the data the tool encodes, while the tools effectiveness describes how much design choices are made in deference to perceptual saliency \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a, munznerVisualizationAnalysisDesign2014}. Mackinlay's definition of expressiveness is formalized at the visual encoding level, which as shown in figure~\ref{fig:intro_retinal_variables} refers to the components of a graphic such as the color or position of a glyph. Bertin first classified these graphic components as retinal variables and discussed which types of data they can express \cite{bertinIIPropertiesGraphic2011} and how they are composited on point, line, and area graphical marks, as shown in figure~\ref{fig:intro_retinal_variables} correspond. Marks can be generalized to glyphs, which are graphical objects that convey one or more attributes of the data entity mapped to it\cite{ware2019information, munznerVisualizationAnalysisDesign2014}. Mackinlay's expressiveness criteria is well defined for the visual variables, such that he suggests the viability of a strict encoding relation that is a homomorphic mapping which preserves some binary operator from one domain to another \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987}. We expand on this suggestion by proposing that monoid action equivariance is a strict condition of building valid encoders. Mackinlay does not provide a generalized criteria for plot types, instead embedding the requirements within the definition of the charts.

@@

\begin{figure}[H]
    \includegraphics[width=1\textwidth]{figures/intro/dataset_diagram.png}
    \caption{One way to describe data is by the connectivity of the points in the dataset. A database for example is often discrete unconnected points, while an image is an implicitely connected 2D grid. This image is from the Data Representation chapter of the MayaVi 4.7.2 documentation.\cite{DataRepresentationMayavi}}
    \label{fig:intro_data_format}
\end{figure}

As shown in figure~\ref{fig:intro_data_format}, there are many types of continuity in data. A database typically consists of unconnected records, while an image is an implicit 2D grid and a network is some sort of explicitly connected graph. In this work we will refer to the points of the dataset as \textit{records} to indicate that a point can be a vector. Each \textit{component} of the record is a single object, such as a temperature, a color, or an image. The way in which these records are connected is the \textit{connectivity}, \textit{continuity}, or more generally \textit{topology}. 
\begin{figure}[H]
    \includegraphics[width=\textwidth]{figures/intro/munzner_datatypes}
   \caption{Keys are unique lookup values used to find individual observations in the dataset. Keys are positional references, and can be coordinates on a map or unique values such as a primary key in a database or a (time, latitude, longitude) index in a data cube. Image modified from a diagram from Munzner's \textit{Visualization Analysis and Design} \cite{munznerVisualizationAnalysisDesign2014}}
   \label{fig:intro_munzner_datatypes}
   \end{figure}

>>>>

As described by Mackinlay, a visualization tool produces a graphical design and an image rendered based on that design. He defines the graphical design as the set encoding relations from data to visual representation\cite{mackinlayAutomatingDesignGraphical1986}, and the design rendered in an idealized abstract space is what throughout this paper we will refer to as a graphic. Mackinlay  proposes that a visualization tool's expressiveness is a measure of how much of the structure of the data the tool encodes, while the tools effectiveness describes how much design choices are made in deference to perceptual saliency \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a, munznerVisualizationAnalysisDesign2014}. Mackinlay's definition of expressiveness is formalized at the visual encoding level, which as shown in figure~\ref{fig:intro_retinal_variables} refers to the components of a graphic such as the color or position of a glyph. Bertin first classified these graphic components as retinal variables and discussed which types of data they can express \cite{bertinIIPropertiesGraphic2011} and how they are composited on point, line, and area graphical marks, as shown in figure~\ref{fig:intro_retinal_variables} correspond. Marks can be generalized to glyphs, which are graphical objects that convey one or more attributes of the data entity mapped to it\cite{ware2019information, munznerVisualizationAnalysisDesign2014}.

Mackinlay's expressiveness criteria is well defined for the visual variables, such that he suggests the viability of a strict encoding relation that is a homomorphic mapping which preserves some binary operator from one domain to another \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987}. We expand on this suggestion by proposing that monoid action equivariance is a strict condition of building valid encoders. Mackinlay does not provide a generalized criteria for plot types, instead embedding the requirements within the definition of the charts.

@@

\begin{figure}[H]
    \includegraphics[width=1\textwidth]{figures/intro/dataset_diagram.png}
    \caption{One way to describe data is by the connectivity of the points in the dataset. A database for example is often discrete unconnected points, while an image is an implicitely connected 2D grid. This image is from the Data Representation chapter of the MayaVi 4.7.2 documentation.\cite{DataRepresentationMayavi}}
    \label{fig:intro_data_format}
\end{figure}
As shown in figure~\ref{fig:intro_data_format}, there are many types of continuity in data. A database typically consists of unconnected records, while an image is an implicit 2D grid and a network is some sort of explicitly connected graph.

\begin{mdframed}[roundcorner=10pt, frametitle= definitions, frametitlerule=true, frametitlebackgroundcolor=gray!10]
 In this work we will refer to the points of the dataset as \textit{records} to indicate that a point can be a vector of heterogenous elements. Each \textit{component} of the record is a single object, such as a temperature, a color, or image. We also generalize \textit{component} to mean all objects in the dataset of a given type, such as all temperatures or colors or images. The way in which these records are connected is the \textit{connectivity}, \textit{continuity}, or more generally \textit{topology}. 
\end{mdframed}

\begin{figure}[H]
    \includegraphics[width=\textwidth]{figures/intro/munzner_datatypes}
   \caption{Keys are unique lookup values used to find individual observations in the dataset. Keys are positional references, and can be coordinates on a map or unique values such as a primary key in a database or a (time, latitude, longitude) index in a data cube. Image modified from a diagram from Munzner's \textit{Visualization Analysis and Design} \cite{munznerVisualizationAnalysisDesign2014}}
   \label{fig:intro_munzner_datatypes}
\end{figure}
